Modelling fine-scale dive behavior with hidden Markov models

### 2020-06-25 Report/Sections/tools.tex
Update Report

>>>>

%%%%%%%%%%%% THROWAWAY STUFF FOR LATER HERE %%%%%%%%%%

\iffalse

\section{for simulation results}

\subsection{Model Misspecification}

There are two ways to simulate the fourier modes of the acceleration vector $\bfA_{s,t}$:

\begin{enumerate}
	\item $$\bbE_1\left(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}\right) = \phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} $$

	\item $$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
	\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
	\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
	\end{array}\right\}$$
\end{enumerate}

\subsubsection{$\bbE_1$}

First, look at $\bbE_1$. This comes with an issue in that when $X^*_{s,t}$ changes values, it will take a while for the process to ``converge" long-run mean $\mu_{X^*_{s,t}}$. In particular, if the behavioral process $X^*$ had been in state 1 for a while and then switches to state 2 for a while, then:

\begin{align*}
	\bbE_1\left(\bfA_{s+1,t}|X^*_{s:s+1,t} = 2, X^*_{s,t} = 1\right) &=  \phi \mu_1 + (1-\phi) \mu_2 \\
	%
	\bbE_1\left(\bfA_{s+2,t}|X^*_{s:s+2,t} = 2, X^*_{s,t} = 1\right) &= \phi\left( \phi \mu_1 + (1-\phi) \mu_2\right) + (1-\phi)\mu_2 \\
	&= \phi^2 \mu_1 + (1-\phi^2) \mu_2\\
	%
	\bbE_1\left(\bfA_{s+n,t}|X^*_{s:s+n,t} = 2, X^*_{s,t} = 1\right) &=  \phi^n \mu_1 + (1-\phi^n) \mu_2 \\
\end{align*}

So, if using $\bbE_1$, there is an assumption that the fourier modes of the acceleration vector $\bfA$ linearly ``converge" to the new value $\mu_i$ from $\mu_j$ when $X^*_{s,t}$ changes from $i$ to $j$. An example of this can be seen in figure (\ref{fig:FOVEDBA}).

 \begin{figure}[h!]
 	\centering
 	\includegraphics[height=3in]{../Code/test.png}
 	\caption{The FoVeDBA of simulated data using $\bbE_1$. Note that the FoVeDBA approaches $\approx$ 300 in state 2 from $\approx$ 10 in state 1 with roughly linear convergence.}
 	\label{fig:FOVEDBA}
 \end{figure}

 Note that if the generative model is correlated, but the inference model isn't, then quick jumps between states 1 and 2 can be missed because the chain never ``converges" to the new mean $\mu_i$, as shown in figure (\ref{fig:STATE}).

\begin{figure}[h!]
	\centering
	\includegraphics[height=3in]{../Code/test2.png}
	\caption{The probability of being in each behavioral state $X^*_{s,t}$ as estimated by the HHMM. The colors represent the true behavioral state.}
	\label{fig:STATE}
\end{figure}

 \subsubsection{$\bbE_2$}

Clearly, the model has issues decodeing the behavioral state when the model is misspecified. Therefore, if the true generative process does not have correlation when jumping between states, then it is better to use $\bbE_2$:

$$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
\end{array}\right\}$$

This fixes the issue of having to ``converge" to the new value of $\mu_i$ when jumping to state $i$. However, note that $\bbE_2$ depends upon both $X^*_{s,t}$ and $X^*_{s-1,t}$. This breaks the assumption in the forward algorithm that $\bfA_{s,t}$ depends only on $X^*_{s,t}$. One way to deal with this issues is to introduce a new hidden state $Z^*_{s,t} = (X^*_{s-1,t},X^*_{s,t})$. This fixes the issue, but now $Z^*_{s,t}$ can take one of $s^2$ values instead of $s$. This grows the state space of our model and can make inference harder.

 \fi

%Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive. The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly. As a result, the data was split into two-second intervals (100 data points each), and each interval was summarized by the Fourier transform of the ODBA and the velocity at the \textit{end} of the time interval. Velocity was taken at the end of the time interval because the behavior of the whale during the time interval has no impact on the velocity of the whale at the beginning of the time interval. Because velocity is highly correlated with the average acceleration within an interval, the mean was subtracted out of each ODBA time series before the Fourier transform was taken. To reduce the dimension of the ODBA Fourier transform, Fourier coefficients corresponding to frequencies less than or equal to 5 hertz were summed and frequencies above 5 hertz were ignored. One nice property of this process is that it reduces the number of data points by a factor of 100, which drastically speeds up parameter estimation. The optimal length of the time interval over which to take the Fourier transform is a tuning parameter that is difficult to find. An ideal time interval should be long enough give useful Fourier coefficients, but short enough to preserve as much information about the vertical velocity of the animal as possible.

%\subsection{Lag Plot}

%In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum. The results of doing so are shown in figure (\ref{fig:lag}). Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that the velocity exhibits a large degree of autocorrelation. While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.

### 2020-06-25 Report/main.tex
Update Report

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}

@@

% Title, authors, affiliations
\title[]{Title on one or two lines without capitals, except after a colon}%\query{Q1}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Association parameters\sep clustered data\sep mean parameters\sep missing data\sep pairwise
likelihood\sep repeated measurements.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62???\sep secondary 62???}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
Insert your abstract here; it should
typically be up to ten lines long. Avoid symbols as much as possible. Formulas
are strongly discouraged, and citations should be avoided. 
The title and the abstract should be concise and descriptive;
list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.

The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling. In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. This results in a vast amount of data which often exhibits many different simultaneous behavioral processes occurring at different time scales.

One solution to this issue is to use a hierarchical hidden Markov model (HHMM). HHMMs model the entire system as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process. One nice property of HHMMs is that its likelihood is relatively easy to compute, facilitating fast maximum likelihood estimates for its associated parameters.

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by a traditional HMMs. To address this issue, it is possible to model small-scale animal behavior as the solution to some stochastic differential equation, but these methods tend to be computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC).

This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. We bridge the gap between discrete hidden Markov models and continuous-time stochastic process models by showing that the two are equivalent under certain conditions. In addition, we extract features from highly structured sub-dive behaviors using signal processing techniques. These features otherwise could not be modeled with a simple HMM. Finally, we apply our method to dive data collected from two Northern resident killer whales off the coast of British Columbia, Canada.

@@

%\input Sections/intro.tex
\input Sections/background.tex
\input Sections/carhmm_ou.tex
\input Sections/model.tex
\input Sections/simulation.tex
\input Sections/results.tex
\input Sections/conclusion.tex

\newpage

@@

\newpage

>>>>

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}

@@

% Title, authors, affiliations
\title[]{Modelling fine-scale dive behavior with hidden Markov models }%\query{Q1}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Association parameters\sep clustered data\sep mean parameters\sep missing data\sep pairwise
likelihood\sep repeated measurements.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62???\sep secondary 62???}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
%Insert your abstract here; it should
%typically be up to ten lines long. Avoid symbols as much as possible. Formulas
%are strongly discouraged, and citations should be avoided. 
%The title and the abstract should be concise and descriptive;
%list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
%http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.

Recent advances in high-frequency tagging
technology have made available a vast
amount of animal movement data. This
rich new data can exhibit simultaneous behavioural processes occurring at different
time scales. One can model these processes
through a hierarchical hidden Markov model
(HHMM), where the system is modelled as a
nested structure of hidden Markov models.
At very short time scales, however, observations can exhibit complicated dependence
structures that cannot be easily captured by
traditional HMMs. We demonstrate how
to incorporate fine-scale processes into the
larger structure of HHMMs while maintaining computational efficiency. We apply our
method to dive and accelerometer data collected from northern resident killer whales off
the coast of British Columbia, Canada.

@@

%\input Sections/intro.tex
\input Sections/background.tex
\input Sections/carhmm_ou.tex
\input Sections/model.tex
\input Sections/simulation.tex
%\input Sections/results.tex
%\input Sections/conclusion.tex

\clearpage

@@

\clearpage
